---
title: "Lab 10: Distances and the Border Zone"

subtitle: 'Ecosystem Science and Sustainability 330'

author:
  - name: Neva Morgan
    email: neva.morgan@colostate.edu
---

### **Objectives**

1.  In this lab we will explore the properties of sf, sfc, and sfg features & objects; how they are stored; and issues related to distance calculation and coordinate transformation.

2.  We will continue to build on our data wrangling and data visualization skills; as well as document preparation via Quarto and GitHub.

#### Loading the libraries:

```{r}
# spatial data science
library(tidyverse)
library(sf)
library(units)

# Visualization
library(gghighlight)
library(ggrepel)
library(knitr)

#Loaded AOI later with Question 1.2!
```

## **Question 1:**

For this lab we need three (3) datasets.

Spatial boundaries of continental USA states (1.1) Boundaries of Canada, Mexico and the United States (1.2) All USA cites (1.3)

### **1.1 Define a Projection**

```{r}
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'
```

[Meaning of this beginning code:]{.underline}

+proj=eqdc → Equidistant Conic projection

+lat_0=40 → Latitude of the projection’s center (40°N)

+lon_0=-96 → Central meridian (96°W)

+lat_1=20 → First standard parallel (20°N)

+lat_2=60 → Second standard parallel (60°N)

+x_0=0 → False easting (0 meters)

+y_0=0 → False northing (0 meters)

+datum=NAD83 → Uses the North American Datum 1983 (NAD83)

+units=m → Units are in meters

+no_defs → No additional default parameters from PROJ’s database

### **1.2 - Get USA state boundaries**

```{r}
remotes::install_github("mikejohnson51/AOI")

library(AOI)

remotes::install_github("ropensci/USAboundaries")
remotes::install_github("ropensci/USAboundariesData")
remotes::install_github("ropenscilabs/rnaturalearthdata")

# Getting USA Boundaries:

us <- aoi_get(state = 'conus') |>
  glimpse()


# Setting it to the projection we made earlier:

us_eqdc <- st_transform(us, crs = eqdc)

st_crs(us_eqdc)
```

Based on the glimpse of what the us boundary contains from the AOI , we can see that this data needed to be changed to reflex the earlier rpojection we made for mapping in North America. By using st_transform and st_crs, I was able to form the data, to reflect the boundaries we want it to be projected into!

### **1.3 - Get country boundaries for Mexico, the United States of America, and Canada**

```{r}

bound <- aoi_get(country = c("MX", "CA", "USA"))


```
